---
title: "scRNASeq CHAT Analysis Details"
author: "The CHAT Group"
date: "1/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      results = FALSE, 
                      message = FALSE, 
                      warning = FALSE)
```

## Intro

This repository contains the code and data that were used to perform the 
analysis of human scRNASeq data from patients undergoing VA-ECLS in support of 
our paper describing the presence and function of human CD4+ cells expressing 
CHAT. Using the code and data provided here, the panels of Figure 1 of that 
paper can be reproduced.

Please note that this repo contains large data files required to reproduce key
elements of the our CHAT cell analysis. If you do not have Git LFS installed,
you will only get the soft links to these files, not the files themselves, when
you clone the repo. That is fine, unless you want to reproduce the analysis (as
opposed to just reviewing the code).

Luckily, LFS support is easy to install. See here: https://git-lfs.github.com

## Dataset and Cell Classification

Alra imputation (https://github.com/KlugerLab/ALRA) was performed on the
original scRNASeq dataset (GSE127221). But to use the Monocole pipeline, raw 
counts were back computed from the normalized/transformed imputed values (by 
taking the natural log, subtracting 1, and dividing by 10,000). This data is 
in cds_rd.rds in the `final_data` folder. Cells were defined based on 
following definitions.

```{r, eval=FALSE}
library(monocle3)

 # this cds contains reconstituted counts based on ALRA imputation
 # reduce_dimensions has been run on it, but we will run it again below

cds <- readRDS("final_data/cds_rd.rds")
dat.imp <- t(normalized_counts(cds))

   
library(foreach)
cellTypeDefs <- list(
  "B Cells" = list(markers = c("CD19", "CD3"),
                       gate = c(1, 0)),

  # CD4 subpopulations can by CD2+/- and FOXP3 +/-, but not double positive
  # Except for CD4 regulatory T which are double positive
  "CD4+ Naive T" = list(markers = c("CD3", "CD4", "CD8", "CD2", "FOXP3", 
                                    "NCAM1"),
                       gate = c(1, 1, 0, 0, 0, 0)),
  "CD4+ Naive T" = list(markers = c("CD3", "CD4", "CD8", "CD2", "IL2RA", 
                                    "FOXP3", "NCAM1" ),
                       gate = c(1, 1, 0, 0, 0, 1, 0)),
  "CD4+ Memory T" = list(markers = c("CD3", "CD4", "CD8", "CD2", "B3GAT1", 
                                     "FOXP3", "NCAM1"),
                       gate = c(1, 1, 0, 1, 0, 0, 0)),
  "CD4+ Memory T" = list(markers = c("CD3"
                                     , "CD4", "CD8", "CD2", "B3GAT1", "IL2RA", 
                                     "FOXP3", "NCAM1"),
                       gate = c(1, 1, 0, 1, 0, 0, 1, 0)),
  "CD4+ Effector T" = list(markers = c("CD3", "CD4", "CD8", "CD2", "B3GAT1", 
                                       "FOXP3", "NCAM1"),
                       gate = c(1, 1, 0, 1, 1, 0, 0)),
  "CD4+ Effector T" = list(markers = c("CD3", "CD4", "CD8", "CD2", "B3GAT1", 
                                       "IL2RA", "FOXP3", "NCAM1"),
                       gate = c(1, 1, 0, 1, 1, 0, 1, 0)),
  "CD4+ Reg T" = list(markers = c("CD3", "CD4", "CD8", "IL2RA", "FOXP3", 
                                  "NCAM1"),
                       gate = c(1, 1, 0, 1, 1, 0)),
  
  "CD4+ TH1" = list(markers = c("CD3", "CD4", "CD8", "IFNG", "IL4", "CXCR3"),
                       gate = c(1, 1, 0, 1, 0, 1)),
  
  "CD4+ TH2" = list(markers = c("CD3", "CD4", "CD8", "IFNG", "IL4", "PTGDR2", 
                                "CCR4"),
                       gate = c(1, 1, 0, 0, 1, 1, 1)),

  "CD4+ TH17" = list(markers = c("CD3", "CD4", "CD8", "IL17A", "IL17B", "CCR6"),
                       gate = c(1, 1, 0, 1, 1, 1)),

  "CD8+ Memory T" = list(markers = c("CD3", "CD4", "CD8", "CD2", "B3GAT1", 
                                     "NCAM1"),
                       gate = c(1, 0, 1, 1, 0, 0)),
  "CD8+ Naive T" = list(markers = c("CD3", "CD4", "CD8", "CD2", "NCAM1"),
                       gate = c(1, 0, 1, 0, 0)),
  "CD8+ Effector T" = list(markers = c("CD3", "CD4", "CD8", "CD2", "B3GAT1", 
                                       "NCAM1"),
                       gate = c(1, 0, 1, 1, 1, 0)),
  
  "NKT CD4+" = list(markers = c("CD3", "CD4", "CD8", "CD19", "NCAM1"),
                       gate = c(1, 1, 0, 0, 1)),

  "NKT CD8+" = list(markers = c("CD3", "CD4", "CD8", "CD19", "NCAM1"),
                       gate = c(1, 0, 1, 0, 1)),

  "NKT CD4- CD8-" = list(markers = c("CD3", "CD4", "CD8", "CD19", "NCAM1"),
                       gate = c(1, 0, 0, 0, 1)),
  
  "NK" = list(markers = c("CD3", "CD19", "NCAM1"),
                       gate = c(0, 0, 1)),
  
  "Monocytes" = list(markers = c("CD14", "CD2", "CXCR2"),
                       gate = c(1, 1, 1)),
  
  "DC" = list(markers = c("CD3", "CD14", "CD19", "NCAM1", "CEACAM8", "CD1C", 
                          "HLA-DRA"),
                       gate = c(0, 0, 0, 0, 0, 1, 1)),
  
  "Erythrocytes" = list(markers = c("HBA1"),
                       gate = c(7))
)


classify <- function(dat, types) {
  class <- rep(NA, nrow(dat))
  res <- foreach(i = 1:length(types)) %do% {
    m <- types[[i]][["markers"]]
    cl <- foreach(j = 1:length(types[[i]][["markers"]]), .combine = "&") %do% {
      if(types[[i]][["gate"]][j]) {
        if(types[[i]][["markers"]][j] == "CD3") {
          return(dat[ , "CD3D"] > types[[i]][["gate"]][j] - 1 |
                   dat[ , "CD3E"] > types[[i]][["gate"]][j] - 1 | 
                   dat[ , "CD3G"] > types[[i]][["gate"]][j] - 1 ) 
        } else if(types[[i]][["markers"]][j] == "CD8") {
          return(dat[ , "CD8A"] > types[[i]][["gate"]][j] - 1 |
                   dat[ , "CD8B"] > types[[i]][["gate"]][j] - 1 | 
                   dat[ , "CD8B2"] > types[[i]][["gate"]][j] - 1 ) 
        } else {
          return(dat[ , m[j]] > types[[i]][["gate"]][j] - 1) 
        }
      } else {
        if(types[[i]][["markers"]][j] == "CD3") {
          return(dat[ , "CD3D"] == 0 &
                   dat[ , "CD3E"] == 0 )
        } else if(types[[i]][["markers"]][j] == "CD8") {
          return(dat[ , "CD8A"] == 0 &
                   dat[ , "CD8B"] == 0 & 
                   dat[ , "CD8B2"] == 0 ) 
        } else {
          return(dat[ , m[j]] == 0) 
        }
      }
    }
    if(length(which(cl)) > 0) {
      
      class[which(cl & !is.na(class))] <- "AMBIG"
      class[which(cl & is.na(class))] <- names(types)[i]
    }
  }
  
  class
}

types <- classify(dat.imp, cellTypeDefs)
types[ which(types == "AMBIG") ] <- NA
types[ which(is.na(types)) ] <- "Unknown"
colData(cds)$cell_type <- types
saveRDS(cds, "final_data/cds_rd.rds")
```

## Patient selection

We limit our analysis to those patients with more than 10 CD4+ lymphocytes to 
analyze.

```{r}

library(monocle3)

cds <- readRDS("final_data/cds_rd.rds")
ix <- grep("CD4\\+", pData(cds)$cell_type)
cds <- cds[ ,ix]

ix.gt10 <- which(table(pData(cds)$Sample_ID) > 10)
ids.gt10 <- names(table(pData(cds)$Sample_ID))[ix.gt10]
ix.cds <- which(pData(cds)$Sample_ID %in% ids.gt10)
cds <- cds[,ix.cds]
pData(cds)$Sample_ID <- factor(pData(cds)$Sample_ID)
saveRDS(cds, "final_data/cds_rd_gt10.rds")

```

## CHAT Expression levels

First we can look at the distribution of CHAT counts among the CD4+ T 
Lymphocytes. We can visualize this either as a violin plot or a sorted line 
plot.

```{r}
library(ggplot2)
library(monocle3)

cds <- readRDS("final_data/cds_rd_gt10.rds")

gd <- data.frame(counts = as.vector(normalized_counts(cds)["CHAT", ]))
a <- ggplot(gd, aes(x=" ", y=counts)) +
  geom_violin(width=0.4, fill="black") + 
  xlab(" ") + 
  ylab("CD4+ CHAT Expression\n(log normalized counts)") +
  theme_bw()


b <- ggplot(gd, aes(x=1:length(gd$counts), y=sort(counts))) +
  geom_line() + 
  xlab(" ") + 
  ylab("Sorted CD4+ CHAT Expression\n(log normalized counts)") +
  theme_bw()

library(ggpubr)
svg(file="Figures/chat_violin.svg", height=3, width=7)
ggarrange(a + theme(plot.margin = unit(c(1,0,0,1), "cm")),
          b + theme(plot.margin = unit(c(1,0,0,1), "cm")), 
          labels = c("A", "B"))
dev.off()
```

We can also look at expression of some other markers of interest.

```{r}
library(reshape2)

ix.g <- which(rownames(cds) %in% c("CHAT", "CD4", "CD8A", "PPIA", "B2M"))
gd <- normalized_counts(cds)[ix.g, ]
gd <- as.data.frame(t(as.matrix(gd)))
gd_m <- melt(gd)
a <- ggplot(gd_m, aes(x=" ", y=value)) +
  geom_violin(width=0.4, fill="black") + 
  xlab(" ") + 
  ylab("CD4+ CHAT Expression\n(log normalized counts)") +
  facet_wrap( ~ variable) +
  theme_bw()

svg(file="Figures/chat_related_violin.svg", height=5, width=7)
print(a)
dev.off()
```

We can also see how chat varies by patient.

```{r}
library(ggplot2)
library(gridExtra)
library(ggpubr)

# get proportion of cd4 cells that are CHAT positive
cd4 <- table(pData(cds)$Sample_ID)
cd4_chat <- table(pData(cds)$Sample_ID,
                  normalized_counts(cds)["CHAT",]>0)[,'TRUE']
chatp <- cd4_chat/cd4

gd <- data.frame(counts = as.vector(normalized_counts(cds)["CHAT",]),
                 sample = pData(cds)$Sample_ID)
gd$sample <- gsub("Patient_", "", as.character(gd$sample))
gd$sample <- factor(gd$sample, labels=c(1:33))

# plot CHAT expression among CD4+ cells for each patient
p<-ggplot(gd, aes(x=sample, y=counts)) +
  geom_jitter(width=0.1, fill="gray", size=0.2, alpha=0.5) + 
  xlab(" ") + 
  ylab("CD4+ CHAT Expression\n(log normalized counts)") +
  theme_bw()

# add annotation table below plot to cross-check
df <- data.frame(proportion = as.vector(round(chatp,2)), 
                 cd4 = as.vector(cd4),
                 cd4_chat = as.vector(cd4_chat),
                 ID = as.numeric(gsub("Patient_", "", names(cd4))))
df <- df[order(df$ID),]

tab_theme <- gridExtra::ttheme_default(
    core = list(fg_params=list(cex = 0.5, col="#333333"),
            bg_params = list(col="#ffffff", fill="#ffffff", lwd=0)),
    colhead = list(fg_params=list(cex = 0.5)),
    rowhead = list(fg_params=list(cex = 0.5)))

b <- tableGrob(t(df[ ,1:3]), 
                cols=NULL, 
                theme=tab_theme) 
b$widths <- unit(rep(.96*(1/ncol(b)), ncol(b)), "npc")
b$widths[1] <- unit(.04 * (1/ncol(b)) * ncol(b), "npc")

c <- ggpubr::as_ggplot(b)
f <- arrangeGrob(p + theme(plot.margin = unit(c(0.5,.25,0,.25), "cm")), 
             c + theme(plot.margin = unit(c(0,0,0,.2), "cm")), 
             nrow=2,
             heights = c(.8, .2))
svg(file="Figures/chat_bypatient.svg", height=4, width=12)
grid::grid.draw(f)
dev.off()
```

Next we plot the proportion of CHAT+ cells for each patient as single dot 
plot to better visualize the distribution.

```{r, eval=TRUE}

gd <- data.frame(proportion = as.vector(chatp), 
                 sample_id = names(chatp))

low <- gd$sample_id[which(gd$proportion < 0.3)]
med <- gd$sample_id[which(gd$proportion > 0.3 & gd$proportion < 0.62)]
high <- gd$sample_id[which(gd$proportion > 0.62)]

svg(file="Figures/chatp_violin.svg", height=4, width=3)
ggplot(gd, aes(x=" ", y=proportion)) + 
#  geom_violin(width=0.2, fill="#cccccc") + 
  geom_jitter(width=0.1) +
  ylim(0,1) +
  xlab(" ") + 
  ylab("CHAT+ / All CD4+ Lymphocytes") +
  theme_bw()

dev.off()


```

## UMAP Plot

It is sometimes instructive to visualize the scRNASeq data reduced to 2 
dimensions using the UMAP algorithm. Here we produce these plots to see if 
there is clustering of CHAT+ cells among all CD4+ cells, or subpopulations of 
CD4+ cells.

## UMAP

```{r}
cds = align_cds(cds, num_dim = 100, alignment_group = "Sample_ID")
cds <- reduce_dimension(cds, umap.metric = "euclidean")
pData(cds)$CHATP <- normalized_counts(cds)['CHAT',] > 0
pData(cds)$cell_type <- factor(pData(cds)$cell_type)

svg(filename="Figures/umap_type.svg", width = 5.5, height=4)
plot_cells(cds, color_cells_by = "cell_type", cell_size = 1, alpha=0.5, label_cell_groups = FALSE) + 
  scale_color_brewer(palette="Set1")
dev.off()

svg(filename="Figures/umap_chatp.svg", width = 5, height=4)
plot_cells(cds, color_cells_by = "CHATP", cell_size = 1, alpha=0.5, label_cell_groups = FALSE) + 
  scale_color_brewer(palette="Set1")
dev.off()

svg(filename="Figures/umap_chatp_stratified.svg", width = 8, height=7)
plot_cells(cds, color_cells_by = "CHATP", cell_size = 1, alpha=0.5, label_cell_groups = FALSE) + 
  scale_color_brewer(palette="Set1") + 
  facet_wrap(~ cell_type + CHATP)
dev.off()

```

## Survival analysis

Here we determine whether the proportion of CHAT+ cells among all CD4+ cells 
is related to survival, using the median proportion as the cutoff for the 
Kaplan Meijer analysis.

```{r}
library(survival)
library(survminer)

# load survival data
md <- readRDS("final_data/metadata.rds")

ix <- match(md$Sample_ID, names(chatp))
md$cellp <- chatp[ix]

# censor survival at 30 days
md$surv_time2 <- md$surv_time
md$surv_time2[which(md$surv_time2 > 720)] <- 720
surv <- Surv(time = md$surv_time2, event = md$surv_time2 < 720)
md$surv <- surv


# binarize proportion of CD4+ cells that are ChAT+ using median as cutoff.
md$high_count <- factor(md$cellp > median(md$cellp, na.rm=TRUE), 
                        labels=c("Low", "High"))
#survdiff(surv ~ high_count, data = md)

# And plot the Kaplan Meier curves along with log-rank test p-value
svg(filename = "Figures/chat_p_30day_survival.svg", width=5, height=3)
p <- ggsurvplot(survfit(surv ~ high_count, data=md), 
           color = "black", 
           ggtheme = theme(legend.title = element_text(size=10, face="bold"),
                           legend.text = element_text(size=10)), 
           data=md, 
           pval=TRUE,
           pval.size = 3.5,
           linetype = "strata",
           risk.table = TRUE,
           legend.labs = c("Low", "High") )
print(p)
dev.off()

```

